---
title: "Heart and Estrogen/progestin Replacement Study (HERS)"
author: "Symon Kimitei"
date: "October 31, 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
DATA WRANGLING AND DATA VISUALIZATION IN RMARKDOWN
Data:  “The Heart and Estrogen/progestin Replacement Study (HERS) is a randomized, double-blind, 
placebo-controlled trial designed to test the efficacy and safety of estrogen plus progestin therapy 
for prevention of recurrent coronary heart disease (CHD) events in women. The participants are postmenopausal 
women with a uterus and with CHD as evidenced by prior myocardial infarction, coronary artery bypass graft surgery, 
percutaneous transluminal coronary angioplasty, or other mechanical revascularization or at least 50% occlusion of a 
major coronary artery. Between February 1993 and September 1994, 20 HERS centers recruited and randomized 2763 women”. 

A simple random sample of the data is provided with the quantitative variables in the file “HW2QuantVar.xlsx” and 
the categorical variables in the file “HW2CatVar.xlsx”.
	
Specify path, load all the necessary libraries and import the necessary excel files.
Merge the HW2QuantVar.xlsx and HW2CatVar.xlsx data sets by "id" into one tibble and
Save the tibble in your R session as the object "d1full".
```{r}
#1 Specify file path
setwd("C:/Users/skimitei/Desktop/r_code")

library(dplyr)
library(tidyverse)
library(readxl)
library(ggplot2)
library("ggpubr")

# Import HW2QuantVars.xlsx file
HW2QuantVars <- read_excel("HW2QuantVars.xlsx")
head(HW2QuantVars)
dim(HW2QuantVars) # Dimension: 1000 rows by 23 columns

# Import HW2QuantVars.xlsx file
HW2CatVars <- read_excel("HW2CatVars.xlsx")
head(HW2QuantVars )
dim(HW2CatVars) # Dimension: 1000 rows by 16 columns



#2 Merge the two files into one
d1full <- merge(HW2QuantVars, HW2CatVars, by="id", all = TRUE)
d1full<- d1full %>%
    select(-matches("id"))

# 3.	Use R to answer the following:
# a.	What are the dimensions of "d1full"? 1000 by 38
# b.	How many subjects do you have in your sample?   1000 subjects
# c.	How many variables are in your sample?  38 variables including the ID variable

#d1full<-na.omit(d1full)
head(d1full)
dim(d1full) # New dimension is 1000 rows(subjects) by 38 columns(variables) including the id column
write.csv(d1full, file = "d1full.csv")



```


4.	Create a barchart and frequency table for the variable "diabetes".  What percent of your sample are diabetic
```{r}
d1full %>% 
    group_by(diabetes) %>% 
    summarize(n = n(), na.rm = TRUE) %>%
    ggplot(aes(x = diabetes, y = n, fill=diabetes)) + 
    geom_col() +geom_bar(stat="identity") #+ coord_flip() # flip the two axes

#Create the frequency table for the variable "diabetes"
d1full %>% 
    group_by(diabetes) %>% 
    summarise(n = n()) %>%
    mutate(freq = n / sum(n))%>%
    mutate(percent_freq = paste0(round(100 * freq, 0), "%"))

```


5.	Subset variables: age, BMI, drinkany, exercise, and glucose for only non-diabetics.  
Next create a new tibble called d1 which has only age, BMI, drinkany, exercise, and glucose for non-diabetics.  
```{r}
d1<-d1full %>% 
  filter(diabetes=="no") %>%
  select(age,BMI,drinkany,exercise,glucose,diabetes) 
  # Add the diabetes column to show that all the elements of this vector all all of value "no"

head(d1)
dim(d1)
```


6.	Gain a general understanding of the variables in your data set by calculating summary statistics for your variables.  Combine results for quantitative variables into one table or tibble.

```{r}
# Gain a general understanding of all the variables in d1full
d1full %>% 
  summary()

```

```{r}
#=======================================================
# Subset the dataframe d1full so that it only contains quantitative variables
d1full %>% 
  select_if(is.numeric)%>%  # select_if subsets conditionally
  summary()  # display the summary statistics for the subsetted quantitative variables

# Store the subsetted variables into a new tibble
num_d1full<-d1full %>% 
  select(DBP,SBP,age,age10,glucose,weight,medcond)
head(num_d1full)
```


7.	For regression, we need to include only complete data for analysis.  Omit na's now to prepare for regression.

```{r}
# Store the subsetted variables into one tibble
num_d1full<-d1full %>% 
  na.omit() %>%
  select(DBP,SBP,age,age10,glucose,weight,medcond)

# View the first 6 observations
head(num_d1full)

# View the summary statistics to ensure that there are no NA values
summary(num_d1full)
   
```
8.	Create a plot to visualize the relationship between glucose and BMI (for non-diabetics) using ggplot.  Treat glucose as the outcome variable and BMI as the predictor variable.  Add a regression line to the scatterplot and make at least 2 customizations to your plot using options illustrated in class, on the ggplot "cheatsheet" from RStudio, or the internet.

*Review & correct this. 
```{r}
d1full$BMI<-as.numeric(as.character(d1full$BMI)) # Convert BMI to numeric datatype


d1full %>%
    filter(diabetes=="no")%>%
    select(BMI,glucose,exercise) %>%
    ggplot(aes(x=BMI,y=glucose,color=exercise,na.rm = TRUE))+  # color the points by exercise
    geom_point(size=2,alpha=0.7)+geom_smooth(method = "lm",se = FALSE, color = "blue")  # Display the regression line in blue


    

  
```


9.	Calculate the correlation between glucose and BMI (for non-diabetics).
```{r}

# Measure the correlation between glucose and BMI
d1full$BMI<-as.numeric(as.character(d1full$BMI)) # Convert BMI to numeric datatype
d1full$glucose<-as.numeric(as.character(d1full$glucose)) # Convert glucose to numeric datatype
d1full %>%
  subset(diabetes == "no") %$%
  cor.test(glucose, BMI) # 0.2567985  which indicates a weak degree of linear dependence between 
                         # glucose and BMI for non-diabetics
  
```

10.	Fit a model predicting glucose by exercise for non-diabetics.  Run diagnostics on the model.  Write 2-4 sentences reporting and interpreting the output.
a.	Note: You may need to recode your yes/no categorical variables to 0/1 indicators.  If so, make sure there is code to show how you do this in R. 
b.	Show the code used to fit the model and create diagnostic output.
c.	Specify the model.
d.	Report the R2 value
e.	Is the model valid?  (check assumptions)
f.	Make one prediction using R.  Interpret.

```{r}
# Create a new column in the dataset d1full and name it exer_yes_no which assigns 1 if exercise is "yes" and 0 if exercise is "no"
d1full$exer_yes_no[d1full$exercise=="yes"]<-1
d1full$exer_yes_no[d1full$exercise=="no"]<-0
head(d1full)

# Create a Scatter plot for x vs y
d1full %>%
    select(glucose,exer_yes_no, exercise) %>%
    ggplot(aes(x=exer_yes_no,y=glucose,color=exercise,na.rm = TRUE))+  
     # color the points by exercise
    geom_point(size=2,alpha=0.7)+geom_smooth(method = "lm", color="red") +
    geom_smooth(se = FALSE, color = "red")  # Display the regression line in red

# Fit a linear regression predicting glucose by exercise for non-diabetics
linearMod <- lm(glucose ~exer_yes_no, data=d1full, color="red")
summary(linearMod)

```

f.	Make one prediction using R.  Interpret.
Create a linear regression model in order to establish the relationship between the predictor(x) and response(x) in the form of a mathematical formula.
```{r}

# Make a model prediction in R

modelSummary <- summary(linearMod)
print("model coefficients")
modelCoeffs <- modelSummary$coefficients
modelCoeffs
b_1 <- modelCoeffs["exer_yes_no", "Estimate"]
b_1
print("y intercept")
b_0 <- modelCoeffs["(Intercept)", "Estimate"]
b_0

# Make a glucose prediction for the case when exercise is "yes"

print("predicted glucose level when exercise is yes ")
yes_no=1
pred_glucose<-b_0+b_1*yes_no
print(pred_glucose)

```



11.	Fit a predicting glucose among non-diabetics by exercise, age, drinkany, and BMI. Write 2-4 sentences reporting and interpreting the output.  Report whether this model performs better than the model in #10.
a.	Note: You may need to recode your yes/no categorical variables to 0/1 indicators.  If so, make sure there is code to show how you do this in R. 
b.	Show the code used to fit the model and create diagnostic output.
c.	Specify the model.
d.	Report the R2 value
e.	Is the model valid?  (check assumptions)

```{r}
d1full$drink_yes_no[d1full$drinkany=="yes"]<-1
d1full$drink_yes_no[d1full$drinkany=="no"]<-0
head(d1full)
fit2 <- lm(glucose ~ exer_yes_no + age + drink_yes_no + BMI, data=d1full)
summary(fit2)
```
```{r}
# capture model summary as an object
modelSummary <- summary(fit2)  


print("model coefficients")
modelCoeffs <- modelSummary$coefficients
modelCoeffs

print("y intercept")
intercept.estimate0 <- modelCoeffs["(Intercept)", "Estimate"]
intercept.estimate0

# If the beta coefficient is positive, the interpretation is that for every 1-unit 
# increase in the predictor variable, the outcome variable will increase by the beta coefficient 
# value and vice versa
print("beta estimate for exer_yes_no variable")
beta.estimate1 <- modelCoeffs["exer_yes_no", "Estimate"]
beta.estimate1

print("beta estimate for age variable")
beta.estimate2 <- modelCoeffs["age", "Estimate"]
beta.estimate2

print("beta estimate for BMI variable")
beta.estimate5 <- modelCoeffs["BMI", "Estimate"]
beta.estimate5

print("f_statistic")
f_statistic <- modelSummary$fstatistic
f_statistic
```





